Type-Safe Modular Parsing

### 2017-01-13 paper/sections/Introduction.tex
Progress on Intro

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{}. Modern Programming Languages enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{}. The
Expression Problem refers to the difficulty of writing data
abstractions that can be easily extended with both new operations and
new data variants. Traditionally the kinds of data abstraction found
in functional languages can be extended with new operations, but
adding new data variants is difficult. The traditional object-oriented
approach to data abstraction facilitates adding new data variants
(classes), while adding new operations is more difficult.

A reason why a solution to the Expression Problem is important in
practice is that it is necessary for the development of
\emph{Software-Product Lines} (SPLs)~\cite{}. A software-product line
is a reusable set of components, which can be combined in multiple ways
to obtain different programs. Programming languages offer a concrete
example for SPLs. A SPL for programming languages would allow us to
model various typical operations of programming languages (such as
evaluation, compilation, or parsing) for various different language
constructs (such as binding, arithmetic, conditional or loops)
independently and separately. For example, evaluation components could be defined
independently for binding and arithmetic constructs. If the language
to be implemented is the pure lambda calculus, only evaluation of
binding constructs is necessary. However, more realistic programming
languages will include arithmetic constructs, and will require 
evaluation for such constructs as well. In this case 
both the component for evaluation of binders and arithmetic 
expressions can be combined to implement the desired functionality.

@@

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use. 
Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. As Kastner~\cite{} notes, the typical
drawback of such techniques is that
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Furthermore they enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{},
virtual classes~\cite{}, as 
well as several other techniques for solving the Expression Problem~\cite{}. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.}

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the perceived need
for sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras are one such technique, which 
Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

@@

>>>>

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{McIlroy68}. Modern Programming Languages enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{wadler1998expression}. The
Expression Problem refers to the difficulty of writing data
abstractions that can be easily extended with both new operations and
new data variants. Traditionally the kinds of data abstraction found
in functional languages can be extended with new operations, but
adding new data variants is difficult. The traditional object-oriented
approach to data abstraction facilitates adding new data variants
(classes), while adding new operations is more difficult.

\begin{comment}
A reason why a solution to the Expression Problem is important in
practice is that it is necessary for the development of
\emph{Software-Product Lines} (SPLs)~\cite{}. A software-product line
is a reusable set of components, which can be combined in multiple ways
to obtain different programs. Programming languages offer a concrete
example for SPLs. A SPL for programming languages would allow us to
model various typical operations of programming languages (such as
evaluation, compilation, or parsing) for various different language
constructs (such as binding, arithmetic, conditional or loops)
independently and separately. For example, evaluation components could be defined
independently for binding and arithmetic constructs. If the language
to be implemented is the pure lambda calculus, only evaluation of
binding constructs is necessary. However, more realistic programming
languages will include arithmetic constructs, and will require 
evaluation for such constructs as well. In this case 
both the component for evaluation of binders and arithmetic 
expressions can be combined to implement the desired functionality.
\end{comment}

@@

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use. 
Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. As Kastner et. al~\cite{Kastner11road} note, the typical
drawback of such techniques is that
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

Syntactic modularization techniques have also been applied to the
problem of \emph{extensible parsing}. There are several approaches
that enable the development of \emph{syntactically} modular parsers or
grammars. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. For instance, \textit{Rats!}~\cite{Grimm2006} has
its own module system for the collection of grammars.  Extensible
compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done ultimately resorting to standard (non-modular) parser
generators. Various techniques supporting languages that can extend
their own syntax, such as SugarJ~\cite{Erdweg2011}, also offer a form
of extensible parsing. However these syntactic approaches do not
support separate compilation and/or modular type-checking of parsing
code either.

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Modular type-checking can report errors 
earlier and in terms of the modular code programmers have written 
in the first place. Separate compilation avoids global compilation
steps, which can be very costly. Furthermore semantic modularization 
enables the composition of compiled binaries as well as ensuring the 
type-safety of the code composed of multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{ernst01FP},
virtual classes~\cite{Ernst:2006}, as 
well as several other techniques for solving the Expression
Problem~\cite{Oliveira:2012}\bruno{more here; yanlin; Torgersen; Odersky...}.

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the perceived need
for sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras~\cite{Oliveira:2012} are one such technique, which 
Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

@@

\item {\bf Algorithmic challenges:} answer Packrat parsing

\item {\bf Typing and Reusability Challenges:} answer Multiple
  inheritance and Object Algebras.

### 2017-01-13 paper/sections/RelatedWork.tex
Progress on Intro

\paragraph{Syntactically Extensible Parsing} Extensible parsing is
used in many different areas. Extensible parser generators are a
mainstream area specially designed for modular syntax and
parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new modules where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars.  NOA~\cite{Gouseti2014} uses Java
annotation processing to gather information about grammars in Java
code. After all information is collected NOA merges all grammar
components, produce an ANTLR~\cite{} grammar, and generates
corresponding parsing code from the ANTLR grammar.  Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In constrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking.

>>>>

\paragraph{Syntactically Extensible Parsing} Extensible parsing is
used in many different areas. Extensible parser generators are a
mainstream area specially designed for modular syntax and
parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new modules where new
non-terminals and production rules can be introduced, some can even
override existing rules in the old grammar modules. For instance,
\textit{Rats!}~\cite{Grimm2006} constructs its own module system for
the collection of grammars.  NOA~\cite{Gouseti2014} uses Java
annotation processing to gather information about grammars in Java
code. After all information is collected NOA merges all grammar
components, produce an ANTLR~\cite{antlr1995} grammar, and generates
corresponding parsing code from the ANTLR grammar.  Those parser
generators focus on the \textit{syntactic extensibility} of grammars:
they rely on whole compilation to generate a global parser, even if
there is only a slight modification in the grammar. Some of those
parser generators may statically check the correctness and unambiguity
of grammars. In constrast, because our approach is based on parser
combinators, there is no support for ambiguity checking.  However, as
far as we are aware, no extensible parser generators support separate
compilation or modular type-checking.
